### 2024-07-05 parts/prelim.tex
starving the beast

\AP Because the functions of interest in this paper have output in $\Nat$ or
$\Rel$, we will only provide the definition of \kl{polyregular functions} for
these two output semigroups, and we refer the reader to \cite{BOKL19} for the
general definition of \kl{polyregular functions}. As a model of computation, we
will use the following definition of transducers, that is based on the marble
transducers \cite{EHB99}, and follows the general pattern of transducers
introduced in \cite{CDTL23}. Note that we only consider $\Rel$-valued (resp.
$\Nat$-valued) functions, but could consider other commutative semigroups.

\AP In order to prove \cref{n-poly-combinatorics:lem}, we will rely on an
equivalent definition of \kl{$\Nat$-polyregular functions} that is more
combinatorial in nature, and is similar in shape to the \emph{finite counting
automata} introduced by \cite{SCHU62}.

\begin{lemma}[$\Rel$-polyregular functions {\cite{CDTL23}}]
    \label{nat-rel-poly:def}
    Let $M$ be a finite monoid, $\mu \colon \Sigma^* \to M$
    be a morphism, $k \in \Nat$, and 
    $\pi \colon M^k \to \Rel$ be a function.
    The following function 
    $\pi^\dagger \colon \Sigma^* \to \Rel$
    is \kl{$\Rel$-polyregular}:
    \begin{equation*}
        \pi^\dagger (w) \defined
        \sum_{w = u_1 \cdots u_k} \pi(\mu(u_1), \dots, \mu(u_k))
        \quad .
    \end{equation*}
    When $\pi$ has co-domain $\Nat$, the function $\pi^\dagger$
    is furthermore \kl{$\Nat$-polyregular}.
    We call $\pi$ the \emph{production function} of $\pi^\dagger$.
    If the monoid $M$ is \kl(monoid){aperiodic}
    then
    the function $f$ is \kl{star-free $\Rel$-polyregular}
    (resp. \kl{star-free $\Nat$-polyregular}).
\end{lemma}

@@

\begin{definition}
    Let $\mathcal{H}$ be a family of functions
    from $\Sigma^*$ to $\Rel$.
    An \intro{$\mathcal{H}$-transducer} $\aTransd$ is
    a tuple $(Q, q_0, \delta, \lambda, F)$ where
         $Q$ is a finite set of states,
         $q_0 \in Q$ is called the initial state of the transducer,
         $\delta \colon Q \times \Sigma \to Q$
            is called the deterministic transition function of the transducer,
         $\lambda \colon Q \times \Sigma \to \mathcal{H}$
            is called the correction function of the transducer,
         and $F \colon Q \to \Rel$ is called the output function of the transducer.
\end{definition}

Based on this syntax, let us now define the semantics of a transducer, in terms
of the function from $\Sigma^*$ to $\Rel$ that it computes. The intuition is
that it suffices to apply the correction function $\lambda$ on every transition
of the underlying automaton, sum the results, and add to this number the output
associated to the final state of the computation.

\begin{definition}
    \label{transducer-sem:def}
    Let $\aTransd \defined (Q,q_0, \delta, \lambda, F)$ be an \kl{$\mathcal{H}$-transducer}.
    The function
    $\aTransd \colon Q \times \Sigma^* \to \Rel$
    is defined inductively as follows:
        $\aTransd(q, \varepsilon) \defined F(q)$, and
        $\aTransd(q, a u) \defined \aTransd(\delta(q,a), u)
            + \lambda(q,a)(u)$.
    The function \intro{computed} by a transducer $\aTransd$
    is $w \mapsto \aTransd(q_0, w)$.
\end{definition}

\begin{example}
    \label{simple-transd:ex}
    Let $\mathsf{Const}$ be the set of constant functions from $\Sigma^*$ to $\Rel$,
    and let us write $\const{n}$ for the constant function equal to $n \in \Rel$.
    In 
    \cref{simple-transd:fig}, we provide a graphical depiction of a
    \kl{$\mathsf{Const}$-transducer} that \kl{computes} the function
    $f \colon \set{a,b}^* \to \Rel$ defined by
    $f(\varepsilon) = 1$,
    $f(aw) = 3 \times \card{w}$,
    and $f(bw) = 3 \times \card{w} + 1$.
\end{example}

\begin{figure}
    \centering
    \begin{tikzpicture}[
        etat/.style={minimum size=2em}
        ]
        \node[etat,state,initial,
            accepting by arrow,
            accepting text={$1$},
            accepting where=below,
            ] (A) at (0,0) {$q_0$};
        \node[etat,state,
            accepting by arrow,
            accepting text={$0$},
            accepting where=below,
            ] (B) at (2.5,0) {$q_1$};

        \draw[->] (A) to[bend right=20] node[midway, below] {$a \mid \const{0}$} (B);
        \draw[->] (A) to[bend left=20] node[midway, above] {$b \mid \const{1}$} (B);

        \draw[->] (B) edge[loop right] node[midway, right]
            {$a,b \mid \const{3}$} (B);
    \end{tikzpicture}
    \caption{A \kl{$\mathsf{Const}$-transducer}, where $\mathsf{Const}$ is the set of constant functions
        from $\set{a,b}^*$ to $\Rel$,
        with a set of states $Q \defined \set{ q_0, q_1 }$,
        an initial state $q_0$,
        an output function $F \colon Q \to \Rel$ such that $F(q_0) = 1$ and $F(q_1) = 0$,
        and correction functions $\lambda(q,a) = \const{0}$, $\lambda(q,b) = \const{1}$,
        and $\lambda(q_1,a) = \lambda(q_1,b) = \const{3}$.
    }
    \label{simple-transd:fig}
\end{figure}

\AP Let us recall that a \intro{counter} in an automaton is a pair $u,w$ of
words together with an integer $n > 1$, such that $\delta(q_0, uw^n) =
\delta(q_0, u)$ and $\delta(q_0, u w^k) \neq \delta(q_0, u)$ for all $k < n$;
where we used the extension of $\delta \colon Q \times \Sigma \to Q$ to words
by defining $\delta(q, \varepsilon) = q$, and $\delta(q,au) =
\delta(\delta(q,a), u)$. An automaton is \reintro{counter-free} if it contains
no \reintro{counters}. This notion can be lifted to
\kl{$\mathcal{H}$-transducers} by considering counters of their underlying
automata.

\AP The notion of \kl{$\mathcal{H}$-transducer} can be used to inductively
define the classes of
\intro{$\Rel $-polyregular functions}
$\intro*\ZPoly[k]$,
\intro{$\Nat $-polyregular functions}
$\intro*\NPoly[k]$,
\intro{star-free $\Rel $-polyregular functions}
$\intro*\ZSF[k]$, and
\intro{star-free $\Nat $-polyregular functions}
$\intro*\NSF[k]$, 
as follows \cite[Theorems 5.15 and 7.10]{DOUE23}: in the base
case, $\reintro*\NPoly[-1] = \reintro*\NSF[-1]  = \reintro*\ZSF[-1] =
\reintro*\ZPoly[-1]$ is the singleton containing the constant function equal to
$0$; and for $k \geq 0$, $\reintro*\ZPoly[k]$ is the set of functions
\kl{computed} by a \kl{$\reintro*\ZPoly[k-1]$-transducer}, and
$\reintro*\NPoly[k]$ is the set of functions \kl{computed} by an
\kl{$\reintro*\NPoly[k-1]$-transducer} \emph{with an $\Nat$-valued output
function $F$}, and similarly for $\reintro*\NSF[k]$ and $\reintro*\ZSF[k]$,
where the transducer is furthermore required to be \kl{counter-free}. Then,
$\reintro*\NPoly = \bigcup_{k \in \Nat} \reintro*\NPoly[k]$, and classes
$\reintro*\NSF$, $\reintro*\ZPoly$ and $\reintro*\ZSF$ are defined analogously.
In the rest of the paper, we will always assume that
\kl{$\reintro*\NPoly[k]$-transducers} and \kl{$\reintro*\NSF[k]$-transducers}
have an $\Nat$-valued output functions, making a clean separation between
$\NPoly$ and $\ZPoly$.

\begin{example}
    \label{non-canonical-transd:ex}
    Let $\BadExOk \colon \set{a}^* \to \Nat$
    be defined as
    $\BadExOk(\varepsilon) = 1$,
    $\BadExOk(a) = 0$,
    and $\BadExOk(a^n) = n-1$ for all $n \geq 2$.
    The function $\BadExOk$ is \kl{computed} by the two \kl{$\NPoly[0]$-transducers}
    depicted in \cref{non-canonical-transd:fig}.
    As a consequence, $f \in \NPoly[1]$.
    Furthermore, because the automaton of 
        \cref{non-canonical-transd:fig:counter-free}
        has no \kl{counter},
    and the correction function only uses functions in $\NSF[0]$,
    $f \in \NSF[1]$.
    \proofref{non-canonical-transd:ex}
\end{example}

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.49\linewidth}
        \begin{tikzpicture}[
            etat/.style={minimum size=2em}
            ]
            \node[etat,state,initial,
                accepting by arrow,
                accepting text={$1$},
                accepting where=below,
                ] (A) at (0,0) {$\varepsilon$};
            \node[etat,state,
                accepting by arrow,
                accepting text={$0$},
                accepting where=below,
                ] (B) at (2,0) {$a$};

            \draw[->] (A) to node[midway, below] {$a \mid 0$} (B);

            \draw[->] (B) to[bend right=45] node[midway, above] {$a \mid 1 - 2 \times \ind{\card{w} = 0}$} (A);
        \end{tikzpicture}
        \caption{This automaton has a \kl{counter}.}
    \end{subfigure}
    \begin{subfigure}[b]{0.49\linewidth}
        \begin{tikzpicture}[
            etat/.style={minimum size=2em}
            ]
            \node[etat,state,initial,
                accepting by arrow,
                accepting text={$1$},
                accepting where=below,
                ] (A) at (0,0) {$\varepsilon$};
            \node[etat,state,
                accepting by arrow,
                accepting text={$0$},
                accepting where=below,
                ] (B) at (2,0) {$a$};

            \draw[->] (A) to node[midway, below] {$a \mid 0$} (B);

            \draw[->] (B) edge[loop right] node[midway, right]
                {$a \mid 1$} (B);
        \end{tikzpicture}
        \caption{This automaton is \kl{counter-free}.}
        \label{non-canonical-transd:fig:counter-free}
    \end{subfigure}
    \caption{Two \kl{$\NPoly[0]$-transducers}
    \kl{computing} the function $\BadExOk$ of \cref{non-canonical-transd:ex},
    that maps $a^n$ to $n$ for all $n \geq 2$, $a$ to $0$, and $\varepsilon$ to $1$.
    }
    \label{non-canonical-transd:fig}
\end{figure}

@@

>>>>

\AP Because the functions of interest in this paper have output in $\Nat$ or
$\Rel$, we will only provide the definition of \kl{polyregular functions} for
these two output semigroups, and we refer the reader to \cite{BOKL19} for the
general definition of \kl{polyregular functions}.

\AP In order to prove \cref{n-poly-combinatorics:lem}, we will rely on an
equivalent definition of \kl{$\Nat$-polyregular functions} that is more
combinatorial in nature, and is similar in shape to the \emph{finite counting
automata} introduced by \cite{SCHU62}. Note that we only consider $\Rel$-valued (resp.
$\Nat$-valued) functions, but could consider other commutative semigroups.

\begin{definition}[$\Rel$-polyregular functions {\cite{CDTL23}}]
    \label{nat-rel-poly:def}
    Let $M$ be a finite monoid, $\mu \colon \Sigma^* \to M$
    be a morphism, $k \in \Nat$, and 
    $\pi \colon M^k \to \Rel$ be a function.
    The following function 
    $\pi^\dagger \colon \Sigma^* \to \Rel$
    is \kl{$\Rel$-polyregular}:
    \begin{equation*}
        \pi^\dagger (w) \defined
        \sum_{w = u_1 \cdots u_k} \pi(\mu(u_1), \dots, \mu(u_k))
        \quad .
    \end{equation*}
    When $\pi$ has co-domain $\Nat$, the function $\pi^\dagger$
    is furthermore \kl{$\Nat$-polyregular}.
    We call $\pi$ the \emph{production function} of $\pi^\dagger$.
    If the monoid $M$ is \kl(monoid){aperiodic}
    then
    the function $f$ is \kl{star-free $\Rel$-polyregular}
    (resp. \kl{star-free $\Nat$-polyregular}).
\end{definition}

@@

@@

\AP The notion of \kl{$\mathcal{H}$-transducer} can be used to inductively
define the classes of \intro{$\Rel $-polyregular functions} $\intro*\ZPoly[k]$,
\intro{$\Nat $-polyregular functions} $\intro*\NPoly[k]$, \intro{star-free
$\Rel $-polyregular functions} $\intro*\ZSF[k]$, and \intro{star-free $\Nat
$-polyregular functions} $\intro*\NSF[k]$, as follows \cite[Theorems 5.15 and
7.10]{DOUE23}: in the base case, $\reintro*\NPoly[-1] = \reintro*\NSF[-1]  =
\reintro*\ZSF[-1] = \reintro*\ZPoly[-1]$ is the singleton containing the
constant function equal to $0$; and for $k \geq 0$, $\reintro*\ZPoly[k]$ is the
set of functions \kl{computed} by a \kl{$\reintro*\ZPoly[k-1]$-transducer}, and
$\reintro*\NPoly[k]$ is the set of functions \kl{computed} by an
\kl{$\reintro*\NPoly[k-1]$-transducer} \emph{with an $\Nat$-valued output
function $F$}, and similarly for $\reintro*\NSF[k]$ and $\reintro*\ZSF[k]$,
where the transducer is furthermore required to be \kl{counter-free}. Then,
$\reintro*\NPoly = \bigcup_{k \in \Nat} \reintro*\NPoly[k]$, and classes
$\reintro*\NSF$, $\reintro*\ZPoly$ and $\reintro*\ZSF$ are defined analogously.
In the rest of the paper, we will always assume that
\kl{$\reintro*\NPoly[k]$-transducers} and \kl{$\reintro*\NSF[k]$-transducers}
have an $\Nat$-valued output functions, making a clean separation between
$\NPoly$ and $\ZPoly$. that \kl{represents} $P$.

Because $f$ has \kl{polynomial
growth}, it is in fact a \kl{$\Nat$-polyregular function} (\cite[Theorem
5.22]{DOUE23}).
\begin{lemma}[CITE]
    \label{polyregular-polynomial-growth:lemma}
    Let $f \in \ZRat$. The following are equivalent:
    \begin{enumerate}
        \item $f \in \ZPoly[k]$.
        \item $f$ has polynomial growth of degree at most $k$.
    \end{enumerate}
    And similarly for $\NPoly$ and $\NRat$.
\end{lemma}

Let us state the crucial compositional properties of
\kl{$\Rel$-polyregular functions} that will be used in this paper
to construct \kl{$\Rel$-polyregular functions}.
\begin{lemma}[CITE]
    \label{stability-polyregular:lemma}
    The class $\NPoly[k]$ (resp. $\ZPoly[k]$) is stable under the operations of
    addition, multiplication, Cauchy product,
    and ``value pinning''.
\end{lemma}

### 2024-07-05 parts/starfree.tex
starving the beast

\begin{lemma}
    \label{decompose-polynomial:lem}
    Let $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
    \kl{$\Rel$-polyregular function},
    where we fix the alphabet $\Sigma = \set{a_1, \dots, a_n}$.
    There exists a computable
    $\omega \in \Nat$,
    and computable 
    polynomials $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^n}$
    such that for all $w \in \Sigma^*$,
    \begin{equation*}
        f\left(a_1^{X_1} \cdots a_n ^{X_n}\right) 
        = P_{\moduloType[\omega](X_1, \dots, X_n)}
        \left(
            \floor{X_1/\omega}, \dots, \floor{X_n/\omega}
        \right)
        \quad .
    \end{equation*}
    \proofref{decompose-polynomial:lem}
\end{lemma}

@@

@@

\begin{theorem}
    \label{zsf-npoly-nsf:thm}
    Let $\Sigma$ be a finite alphabet, 
    and $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
    \kl{$\Nat$-polyregular function}.
    Then, the following are equivalent:
    \begin{enumerate}
        \item $f$ is \kl{ultimately polynomial},
        \item $f \in \ZSF$,
        \item $f \in \NSF$.
    \end{enumerate}
    Furthermore, membership is decidable and conversions are effective.
    \proofref{zsf-npoly-nsf:thm}
\end{theorem}
\begin{proof}
    We proceed by induction on the size of the alphabet $\card{\Sigma}$.

>>>>

\begin{lemma}
    \label{decompose-polynomial:lem}
    Let $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
    \kl{$\Rel$-polyregular function},
    where we fix the alphabet $\Sigma = \set{a_1, \dots, a_n}$.
    There exists a computable
    $\omega \in \Nat_{\geq 1}$,
    and computable 
    polynomials $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^n}$
    such that for all $x_1, \dots, x_n \geq \omega$:
    \begin{equation*}
        f\left(a_1^{x_1} \cdots a_n ^{x_n}\right) 
        = P_{\moduloType[\omega](x_1, \dots, x_n)}
        \left(
            \floor{x_1/\omega}, \dots, \floor{x_n/\omega}
        \right)
        \quad .
    \end{equation*}
    \proofref{decompose-polynomial:lem}
\end{lemma}

@@

\begin{conjecture}
    \label{npoly-zpoly:conjecture}
    Let $k \in \Nat$.
    A function $f \in \ZPoly[k]$
    belongs to $\NPoly[k]$ if and only if
    it is \kl{$(k,\Nat)$-combinatorial}.
    In particular,
    $\NPoly[k] = \ZPoly[k] \cap \NPoly$.
\end{conjecture}

@@

\begin{theorem}
    \label{zsf-npoly-nsf:thm}
    Let $\Sigma$ be a finite alphabet, 
    and $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
    \kl{$\Nat$-polyregular function}.
    Then, the following are equivalent:
    \begin{enumerate}
        \item $f$ is \kl{ultimately polynomial},
        \item $f \in \ZSF$,
        \item $f \in \NSF$.
    \end{enumerate}
    Furthermore, membership is decidable and conversions are effective.
    \proofref{zsf-npoly-nsf:thm}
\end{theorem}
\begin{proof}
    Let $f$ be a \kl{commutative} \kl{$\Nat$-polyregular function}. If $f \in \NSF \subseteq \ZSF$
    then $f$ is \kl{ultimately polynomial} by \cite{CDTL23}.

    % FIXME:
    % -

    We proceed by induction on the size of the alphabet $\card{\Sigma}$.
